```{r}
library(tidyverse)
library(cancensus)
library(knitr)
library(readr)
library(sf)
library(geojsonsf)
library(paletteer)
```

### Loading census using API key

#### First we load our api key and create our vector of columns we would like to extract

```{r}
load("API_KEY.rda")

options(cancensus.api_key = api_key)
options(cancensus.cache_path = "cache")

vectors <- c("v_CA21_1",
             "v_CA21_6",
             "v_CA21_449",
             "v_CA21_1040",
             "v_CA21_1085",
             "v_CA21_905")
```

#### Get the dataset and save to file. Clear api key out of global env. THIS DATA CONTAINS SF INFO FOR CENSUS!!!!

```{r}
census_data <- get_census(
  dataset = "CA21",
  regions = list(
    CSD = "5915022",
    DA = c(
      "59154012",
      "59154105",
      "59154090",
      "59150936",
      "59154101",
      "59154104",
      "59154035",
      "59154103",
      "59154102",
      "59154034",
      "59150945",
      "59154091",
      "59154093",
      "59154099",
      "59150946",
      "59154100",
      "59154078",
      "59154079",
      "59154082",
      "59154081",
      "59154080",
      "59150939",
      "59150938",
      "59154083",
      "59154095",
      "59154084",
      "59150941",
      "59150942",
      "59154085",
      "59154088",
      "59154087",
      "59154089",
      "59154097",
      "59154098",
      "59154096",
      "59154092",
      "59154013",
      "59150952"
    )
  ),
  vectors = vectors,
  labels = "detailed",
  geo_format = "sf",
  level = "DA"
)

can_api_key <- ""
save(census_data, file = "data/census.rda")
```

```{r}
head(census_data)
```

### (note)Then we have to load other data, transform , add sf info, then merge into dataset and add comparison functions

### Loading other datasets , renaming and standardizing columns

```{r}
food_data <- st_read("data/free-and-low-cost-food-programs.shp")  %>%
  select(
    "program_nam",
    "program_sta",
    "meal_cost",
    "local_areas",
    "latitude",
    "longitude",
    "geometry"
  ) %>%
  drop_na("latitude", "longitude") %>%
  # set to wgs 84 as per can census
  st_set_crs(4326)

head(food_data)
head(food_data$geometry)
```

```{r}
crime <- read_csv("data/crime_data_all_neighborhoods.csv")
colnames(crime) <- tolower(make.names(colnames(crime)))

head(crime)
# now we mutate crime data to add shapefile information
# first we create as a utm shapefile due to the data structure,
# then convert to lat/long like census_data and food_data
# crs = 32610 : utm 10
# crs = 4326 : lat/long WGS84
crime_data <- crime %>%
  st_as_sf(coords = c("x", "y"), crs = 32610) %>%
  st_transform(crs = 4326)

# we need to remove (0,0) values for plotting as they cause errors on the axis
# to do that we extract coords then filter where the Y component == 0
crime_coords <- st_coordinates(crime_data)
crime_data <- crime_data[crime_coords[, 2] != 0, ]

head(crime_data)
head(st_geometry(crime_data))
```

### Shapefile distance calculations

```{r}
# example statistics from jonathon
census_geom <- st_geometry(census_data)
census_centroids <- st_centroid(census_geom)

distance_food <- st_distance(census_centroids, food_data)

intersections_food <- st_intersects(census_data, food_data, sparse = FALSE)
intersections_food_500 <- st_is_within_distance(census_data, food_data, sparse = FALSE, dist = 500)
food_contained <- rowSums(intersections_food, dims = 1)
food_within_500 <- rowSums(intersections_food_500, dims = 1)

mean(food_contained)
mean(food_within_500)
```

#### this cell takes a long time to run, idk why right now so avoid running it, all other calcs work independent (it does finish running though)

```{r}
distance_crime <- st_distance(census_centroids, crime_data)
```

```{r}
intersections_crime <- st_intersects(census_data, crime_data, sparse = FALSE)
intersections_crime_500 <- st_is_within_distance(census_data, crime_data, sparse = FALSE, dist = 500)
crime_contained <- rowSums(intersections_crime, dims = 1)
crime_within_500 <- rowSums(intersections_crime_500, dims = 1)


mean(crime_contained)
mean(crime_within_500)
```

### Visualizing our data

```{r}
can_map <- ggplot() +
  geom_sf(data = census_data, color = "blue") +
  theme_minimal() +
  labs(title = "Census data mapping", )

crime_map <- ggplot() +
  geom_sf(
    data = crime_data,
    aes(color = type),
    alpha = 0.8,
    size = 0.5,
    fill = "white"
  ) +
  theme_minimal() +
  labs(title = "Crime data mapping",
       fill = "Legend",
       color = "Crime Type") +
  theme(legend.position = "right") +
  scale_color_paletteer_d("ggthemes::Classic_Purple_Gray_12")

# some extra themes
# scale_color_paletteer_d("ggprism::autumn_leaves")
# scale_color_paletteer_d("beyonce::X70")
# scale_color_paletteer_d("impressionist.colors::te_aa_no_areois")
# scale_color_paletteer_d("ggthemes::Classic_Purple_Gray_12")

food_map <- ggplot() +
  geom_sf(data = food_data, color = "blue") +
  theme_minimal() +
  labs(title = "Food center mapping")

can_map
crime_map
food_map
```

```{r}
crime_food_plot <- ggplot() +
  geom_sf(data = census_data, color = "black") +
  geom_sf(
    data = crime_data,
    aes(color = "geometry"),
    size = 0.4,
    alpha = 0.2
  ) +
  geom_sf(
    data = food_data,
    aes(fill = "yellow"),
    color = "black",
    size = 2,
    alpha = 0.8,
    shape = 21
  ) +
  coord_sf() +
  labs(title = "Free and Low Cost Food Programs plotted against Crime, Vancouver") +
  theme_minimal() +
  theme(legend.position = "right") +
  scale_fill_manual(
    name = "",
    values = "yellow",
    labels = c("Food Programs")
  ) +
  scale_color_manual(name = "",
                     values = "red",
                     labels = c("Crime")) +
  # increase point size for legend only
  guides(fill = guide_legend(override.aes = list(size = 5)),
         color = guide_legend(override.aes = list(size = 5, alpha = 0.5)))

crime_food_plot
```

```{r}
lim_at_plot <- ggplot() +
  geom_sf(
    data = census_data,
    aes(fill = `v_CA21_1040: Prevalence of low income based on the Low-income measure, after tax (LIM-AT) (%)`),
    color = "black",
    size = 0.2
  ) +
  scale_fill_paletteer_c("viridis::viridis", 
                         name = "Low Income (%)",
                         limits = c(0, 100)) +
  labs(title = "Prevalence of low income based on the Low-income measure, after tax") +
  theme_minimal() +
  theme(legend.position = "right")

lim_at_plot 

lico_at_plot <- ggplot() +
  geom_sf(
    data = census_data,
    aes(fill = `v_CA21_1085: Prevalence of low income based on the Low-income cut-offs, after tax (LICO-AT) (%)`),
    color = "black",
    size = 0.2
  ) +
  scale_fill_paletteer_c("viridis::viridis", 
                         name = "Low Income (%)",
                         limits = c(0, 100)) +
  labs(title = "Prevalence of low income based on the Low-income cut-offs, after tax") +
  theme_minimal() +
  theme(legend.position = "right")

lico_at_plot
```

```{r}
lim_crime_food_plot <- ggplot() +
  # lim-at + can map from census data
  geom_sf(
    data = census_data,
    aes(fill = `v_CA21_1040: Prevalence of low income based on the Low-income measure, after tax (LIM-AT) (%)`),
    color = "black",
    size = 0.2
  ) +
  # crime mapping
  geom_sf(
    data = crime_data,
    color = "red",
    size = 0.3,
    alpha = 0.2
  ) +
  geom_sf(
    data = food_data,
    fill = "yellow",
    color = "black",
    size = 2,
    alpha = 0.8,
    shape = 21
  ) +
  # lim-at gradient
  scale_fill_paletteer_c("viridis::viridis", 
                         name = "Low Income (%)",
                         limits = c(0, 100)) +
  labs(title = "Low Income Prevalence + Crime and Food Centers (LIM-AT)") +
  theme_minimal() +
  theme(legend.position = "right")


lim_crime_food_plot                              
```

```{r}
lico_crime_food_plot <- ggplot() +
  # Low income measure (LIM-AT) as a continuous fill
  geom_sf(
    data = census_data,
    aes(fill = `v_CA21_1085: Prevalence of low income based on the Low-income cut-offs, after tax (LICO-AT) (%)`),
    color = "black",
    size = 0.2
  ) +
  # Crime data, set color to red without aes (not mapped)
  geom_sf(
    data = crime_data,
    color = "red",
    size = 0.2,
    alpha = 0.2
  ) +
  # Food programs, set fill to yellow without aes (not mapped)
  geom_sf(
    data = food_data,
    fill = "yellow",   # Fixed yellow fill
    color = "black",
    size = 2,
    alpha = 0.8,
    shape = 21
  ) +
  # Viridis scale for continuous data
  scale_fill_paletteer_c("viridis::viridis", 
                         name = "Low Income (%)",
                         limits = c(0, 100)) +
  labs(title = "Low Income Prevalence + Crime and Food Centers (LICO-AT)") +
  theme_minimal() +
  theme(legend.position = "right")

#TODO: legends for food center


lico_crime_food_plot  
```
