---
output:
  pdf_document: 
    latex_engine: lualatex
  html_document: default
header-includes:
  - \usepackage{graphicx}
---

```{r setup, include=FALSE}
lapply(c("tidyverse", "cancensus", "knitr", "readr", "sf", 
         "geojsonsf", "paletteer", "kableExtra", 
         "broom", "here", "stargazer", "lmtest", "sandwich"), library, character.only = TRUE)
```

# Data Workbook: Group 14 ECON 326

note for grader: if you want to view tables within r studio you need to use knitr and knit to pdf or change format = latex to text

## Members

Sidharth Mirchandani, Radhika Iyer, Hajra Ojha, Anishka Fernandopulle

## Census data

We load previously loaded census data. The code for fetching this data is also shown in this section.

```{r, eval = FALSE}
#| execute: false
load(here("API_KEY.rda"))

options(cancensus.api_key = api_key)
options(cancensus.cache_path = "cache")

vectors <- c("v_CA21_1", "v_CA21_6", "v_CA21_452", "v_CA21_449", "v_CA21_1040", "v_CA21_1085", "v_CA21_905")

region_DA <- c("59154012", "59154105", "59154090", "59150936", "59154101", "59154104", 
               "59154035", "59154103", "59154102", "59154034", "59150945", "59154091", 
               "59154093", "59154099", "59150946", "59154100", "59154078", "59154079", 
               "59154082", "59154081", "59154080", "59150939", "59150938", "59154083", 
               "59154095", "59154084", "59150941", "59150942", "59154085", "59154088", 
               "59154087", "59154089", "59154097", "59154098", "59154096", "59154092", 
               "59154013", "59150952")

census_data <- get_census(
  dataset = "CA21",
  regions = list(CSD = "5915022", DA = region_DA),
  vectors = vectors,
  labels = "detailed",
  geo_format = "sf",
  level = "DA"
)

census_data <- census_data %>%
  mutate(pop_density = `v_CA21_1: Population, 2021` / `Shape Area`)

can_api_key <- ""
save(census_data, file = "../../data/census.rda")
```

```{r table_1}
load(here("data/census.rda"))

n <- nrow(census_data)
kable(head(census_data[, 1:5]), format = "latex", booktabs = TRUE, caption = "Census Data, 2021 (truncated rows and columns)") %>%
  kable_styling(latex_options = c("striped", "hold_position"))
```

## Food data

```{r}
food_data <- st_read(here("data/free-and-low-cost-food-programs.shp"))  %>%
  select(
    "program_nam",
    "program_sta",
    "meal_cost",
    "local_areas",
    "latitude",
    "longitude",
    "geometry"
  ) %>%
  drop_na("latitude", "longitude") %>%
  # set to wgs 84 as per can census
  st_set_crs(4326)
```

```{r table_2}
# Food data processing
food_count <- food_data %>%
  st_set_geometry(NULL) %>% 
  group_by(local_areas) %>%
  summarise(count = n(), .groups = "drop")

food_data_count <- food_data %>%
  left_join(food_count, by = "local_areas") %>%
  distinct(local_areas, .keep_all = TRUE) # one row per neighbourhood

combo_food_census <- census_data %>%
  st_join(food_data_count)

census_data_food <- combo_food_census %>%
  mutate(program_count = replace_na(count, 0),
         food_density = program_count / `Shape Area`)

head(census_data_food[, 1:5]) %>%
  kable(format = "latex", booktabs = TRUE, caption = "Food Data merged with Census Data, 2021 (truncated rows and columns)") %>%
  kable_styling(latex_options = c("striped", "hold_position"))
```

\_

## Crime data

```{r}
crime <- read_csv(here("data/crime_data_all_neighborhoods.csv"), show_col_types = FALSE) %>%
  mutate(TYPE = as_factor(TYPE),
         HUNDRED_BLOCK = as_factor(HUNDRED_BLOCK),
         NEIGHBOURHOOD = as_factor(NEIGHBOURHOOD)) %>%
  filter(!is.na(X) & !is.na(Y))

crime_data <- st_as_sf(crime, coords = c("X", "Y"), crs = "+proj=utm +zone=10") %>%
  st_transform(crs = "+proj=longlat +datum=WGS84")
```

```{r}
intersections <- st_is_within_distance(census_data, crime_data, sparse = FALSE, dist = 5)

crimes_contained <- rowSums(intersections, dims = 1)

census_data_crime <- census_data %>%
  cbind(crimes_contained) %>%
  mutate(crime_density = crimes_contained / Shape.Area)

unique_crimes <- unique(crime_data$TYPE)

for (type in unique_crimes) {
  type_data <- crime_data %>% filter(TYPE == type)
  intersections <- st_is_within_distance(census_data, type_data, sparse = FALSE, dist = 5)
  sum <- rowSums(intersections, dims = 1)
  df <- as.data.frame(sum)
  census_data_crime <- census_data_crime %>% cbind(df$sum) %>% rename_with(~ paste0("crimes_", type), df.sum)
}
```

## Training data

Now we merge all of our data-sets into one table so we can feed it into our model.

```{r}
training_data <- st_join(census_data_crime, census_data_food)

training_data <- training_data %>%
  rename_with(
    ~ gsub(":.*$", "", .),  # Remove everything after the colon, including the colon
    starts_with("v_CA21")
  )

median_lico_at <- median(training_data$v_CA21_1085, na.rm = TRUE)

training_data <- training_data %>%
  mutate(
    low_income = ifelse(
      v_CA21_1085 > median_lico_at, 
      1, 
      0))


training_data[is.na(training_data)] <- 0
```

# Model

## Specification

```{r}
# Specification models
# 1. Model with all variables (including interaction term)
reg_all_vars <- lm(crime_density ~ 
                   food_density:low_income + 
                   food_density + 
                   pop_density.x + 
                   v_CA21_452 +
                   low_income, 
                 data = st_set_geometry(training_data, NULL))

# 2. Model without the interaction term
reg_no_interaction <- lm(crime_density ~ 
                         food_density + 
                         pop_density.x + 
                         v_CA21_452 +
                         low_income, 
                       data = st_set_geometry(training_data, NULL))

# 3. Model with only food density, crime density, and low income
reg_food_crime_low_income <- lm(crime_density ~ 
                                food_density + 
                                low_income, 
                              data = st_set_geometry(training_data, NULL))

# 4. Model with all variables (including interaction term) but replacing v_CA21_452 with v_CA21_449
reg_all_vars_449 <- lm(crime_density ~ 
                       food_density:low_income + 
                       food_density + 
                       pop_density.x + 
                       v_CA21_449 +  # Replaced v_CA21_452 with v_CA21_449
                       low_income, 
                     data = st_set_geometry(training_data, NULL))

models_spec <- list(
  "All Variables" = reg_all_vars,
  "Without Interaction" = reg_no_interaction,
  "Food Density, Crime, Low Income" = reg_food_crime_low_income
)

# Summary for each specification tested
summary_all_vars <- summary(reg_all_vars)
summary_no_interaction <- summary(reg_no_interaction)
summary_food_crime_low_income <- summary(reg_food_crime_low_income)
summary_all_vars_449 <- summary(reg_all_vars_449)


model_summaries <- list(
  "All Variables" = summary_all_vars,
  "Without Interaction" = summary_no_interaction,
  "Food Density, Crime, Low Income" = summary_food_crime_low_income,
  "All Variables (with v_CA21_449)" = summary_all_vars_449
)
```

# RESULTS FROM SPEC

### Function definitions

We needed to manipulate our regression model data into formatted tables that emulated those from the package `stargazer` as we were having significant alignment issues between the data and rendered tables. Guess what.... I was broken and not stargaze4r. lovely overengineering right here though so enjoy

```{r}
all_variables <- unique(c(
  names(coef(reg_all_vars)),
  names(coef(reg_no_interaction)),
  names(coef(reg_food_crime_low_income))
))

all_variables_escaped <- gsub("_", "\\\\_", all_variables)  # Escape underscores on vac23w19o3i21  0i4e09q1wu census variables
all_variables_escaped <- paste0("\\texttt{", all_variables_escaped, "}")  #WRAPFJOERF LATEX

add_stars <- function(estimates, ses, p_values) {
  significance_levels <- ifelse(p_values < 0.01, "***",
                          ifelse(p_values < 0.05, "**",
                          ifelse(p_values < 0.1, "*", "")))
  formatted <- sprintf("%.3f (%.3f)%s", estimates, ses, significance_levels)
  return(formatted)
}

extract_summary <- function(model, all_vars) {
  coefs <- coef(model)
  ses <- sqrt(diag(vcov(model)))
  p_values <- coef(summary(model))[, 4]  
  aligned_coefs <- setNames(rep(NA, length(all_vars)), all_vars)
  aligned_ses <- setNames(rep(NA, length(all_vars)), all_vars)
  aligned_p_values <- setNames(rep(NA, length(all_vars)), all_vars)

  for (var in names(coefs)) {
    aligned_coefs[var] <- coefs[var]
    aligned_ses[var] <- ses[var]
    aligned_p_values[var] <- p_values[var]
  }

  aligned_coefs[is.na(aligned_coefs)] <- 0

  # note bottom footnote stars. order is inverted??????
  significance_levels <- ifelse(aligned_p_values < 0.01, "***",
                          ifelse(aligned_p_values < 0.05, "**",
                          ifelse(aligned_p_values < 0.1, "*", "")))

  # align footnote here
  formatted <- ifelse(is.na(aligned_ses),
                      sprintf("%.3f", aligned_coefs),
                      sprintf("%.3f (%.3f)%s", aligned_coefs, aligned_ses, significance_levels))

  formatted[is.na(aligned_ses)] <- " "

  formatted <- gsub("_", "\\\\_", formatted)
  formatted <- paste0("\\texttt{", formatted, "}")

  return(formatted)
}

extract_model_info <- function(model, covariate_labels, model_title) {
  coefs <- coef(model)
  ses <- sqrt(diag(vcov(model)))
  p_values <- coef(summary(model))[, 4]
  significance_levels <- ifelse(p_values < 0.01, "***",
                                ifelse(p_values < 0.05, "**",
                                       ifelse(p_values < 0.1, "*", "")))
  
  formatted <- sprintf("%.3f (%.3f)%s", coefs, ses, significance_levels)

  result <- data.frame(
    Variable = covariate_labels,
    Estimate = formatted,
    stringsAsFactors = FALSE
  )
  
  result <- rbind(data.frame(Variable = model_title, Estimate = "", stringsAsFactors = FALSE), result)
  
  return(result)
}


r_squared <- c(
  summary(reg_all_vars)$r.squared,
  summary(reg_no_interaction)$r.squared,
  summary(reg_food_crime_low_income)$r.squared
)

f_statistic <- c(
  summary(reg_all_vars)$fstatistic[1],
  summary(reg_no_interaction)$fstatistic[1],
  summary(reg_food_crime_low_income)$fstatistic[1]
)

n_obs <- c(
  length(reg_all_vars$fitted.values),
  length(reg_no_interaction$fitted.values),
  length(reg_food_crime_low_income$fitted.values)
)


model1_aligned <- extract_summary(reg_all_vars, all_variables)
model2_aligned <- extract_summary(reg_no_interaction, all_variables)
model3_aligned <- extract_summary(reg_food_crime_low_income, all_variables)
```

### Data manipulation

```{r}
aligned_table <- data.frame(
  Variable = all_variables_escaped,
  `Model 1` = model1_aligned,
  `Model 2` = model2_aligned,
  `Model 3` = model3_aligned,
  stringsAsFactors = FALSE
)

print(aligned_table)

summary_table <- data.frame(
  Variable = c("R squared", "F statistic", "Observations"),
  `Model 1` = c(sprintf("%.3f", r_squared[1]), sprintf("%.1f", f_statistic[1]), n_obs[1]),
  `Model 2` = c(sprintf("%.3f", r_squared[2]), sprintf("%.1f", f_statistic[2]), n_obs[2]),
  `Model 3` = c(sprintf("%.3f", r_squared[3]), sprintf("%.1f", f_statistic[3]), n_obs[3]),
  stringsAsFactors = FALSE
)

final_table <- rbind(aligned_table, summary_table)
print(final_table)
```

```{r}
covariate_labels <- c(
  "(Intercept)", 
  "Food Program Density: Low Income",
  "Food Program Density",
  "Population Density", 
  "Average Household Size (v\\_CA21\\_452)",
  "Low Income"
)
covariate_labels_2 <- c(
  "(Intercept)",
  "Food Program Density",
  "Population Density", 
  "Average Household Size (v\\_CA21\\_452)",
  "Low Income"
)

covariate_labels_3 <- c(
  "(Intercept)",
  "Food Program Density",
  "Low Income"
)

model_1_table <- extract_model_info(
  model = reg_all_vars,
  covariate_labels = covariate_labels,
  model_title = "Spec 1: All Variables"
)

model_2_table <- extract_model_info(
  model = reg_no_interaction,
  covariate_labels = covariate_labels_2,
  model_title = "Spec 2: Omitted Interaction Term"
)

model_3_table <- extract_model_info(
  model = reg_food_crime_low_income,
  covariate_labels = covariate_labels_3,
  model_title = "Spec 3: Food Density and Low Income"
)
```

### Regression table

```{r, results='asis'}
stargazer(reg_all_vars, reg_no_interaction, reg_food_crime_low_income, type = "latex", 
          covariate.labels = c("Food Program Density", "Population Density", "Household Size", "Low Income", "Food Density:Low Income"),
          dep.var.labels = c("Crime Density")
         )
```

### Model 1:

```{r, results='asis'}
# Model 1: reg_all_vars
stargazer(reg_all_vars, type = "latex", 
          covariate.labels = c("Food Program Density", "Population Density", "Household Size", "Low Income", "Food Density:Low Income"),
          dep.var.labels = c("Crime Density")
         )
```

### Model 2:

```{r, results='asis'}
stargazer(reg_no_interaction, type = "latex", 
          covariate.labels = c("Food Program Density", "Population Density", "Household Size", "Low Income"),
          dep.var.labels = c("Crime Density")
         )
```

### Model 3:

```{r, results='asis'}
# Model 3: reg_food_crime_low_income
stargazer(reg_food_crime_low_income, type = "latex", 
          covariate.labels = c("Food Program Density", "Low Income"),
          dep.var.labels = c("Crime Density")
         )
```

## Robustness

We split the data into sub-groups based on

```{r}
census_data_centroids <- census_data %>%
  mutate(
    centroid = st_centroid(geometry),
    latitude = st_coordinates(centroid)[, 2],
    longitude = st_coordinates(centroid)[, 1]
  ) %>%
  st_set_geometry(NULL)  # Remove geometry for k-means input

# Normalize latitude and longitude for clustering
census_data_normalized <- census_data_centroids %>%
  mutate(
    latitude_scaled = scale(latitude),
    longitude_scaled = scale(longitude)
  )


set.seed(123) # For reproducibility
k <- 4

kmeans_result <- kmeans(census_data_normalized[, c("latitude_scaled", "longitude_scaled")], centers = k)

# Add cluster assignments to the original data
census_data$cluster <- kmeans_result$cluster

# Now, merge the cluster assignments into your training data
training_data_clusters <- st_join(training_data, 
                         census_data %>% select(cluster, geometry), 
                         left = TRUE)
```

```{r}
# Run the model for each cluster
cluster_models <- list()

for (i in 1:k) {
  subset_data <- training_data_clusters %>% filter(cluster == i)
  
  model <- lm(crime_density ~ 
                food_density:low_income + 
                food_density + 
                pop_density.x + 
                v_CA21_452 +
                low_income, 
              data = st_set_geometry(subset_data, NULL))
  
  cat(paste("\nModel Summary for Cluster", i, ":\n"))
  cluster_models[[i]] <- model
  print(summary(model)) # Check the results for each cluster
}
```

```{r, results='asis'}
stargazer(cluster_models, type = "latex", 
          covariate.labels = c("Food Program Density", "Population Density", "Household Size", "Low Income", "Food Density:Low Income"),
          dep.var.labels = c("Crime Density"))
```

```{r}
# Convert the cluster column to a factor
census_data$cluster <- as.factor(census_data$cluster)

# Plot the regions and highlight clusters with adjusted color scale
cluster_plot <- ggplot(data = census_data) +
  geom_sf(aes(fill = cluster), color = "black", lwd = 0.1) +
  scale_fill_viridis_d(name = "Cluster #") +  # Automatically assign colors for clusters
  labs(
       fill = "Cluster #") +
  theme_minimal() +
  theme(legend.position = "right", 
        plot.title = element_text(hjust = 0.5, size = 10, face = "bold"),
        plot.subtitle = element_text(hjust = 0.5, size = 12))

ggsave("census_clusters.png", 
       plot = cluster_plot, 
       width = 10, height = 8, dpi = 300, units = "in")
```

### Heteroskedascity

```{r}
training_data_res <- training_data %>%
  mutate(residuals = resid(reg_all_vars))

# Plot residuals against population density to visually check for heteroskedasticity
ggplot(data = training_data_res, aes(x = pop_density.x, y = residuals)) +
  geom_point(size = 0.2, alpha =0.5, color="blue") +
  labs(x = "Population Density", y = "Residuals") +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5))

```

```{r}
# Extract residuals and calculate squared residuals
training_data_res <- training_data_res %>%
  mutate(residuals_squared = residuals^2)

white_model <- lm(residuals_squared ~ 
                    food_density + pop_density.x + v_CA21_452 + low_income +
                    I(food_density^2) + I(pop_density.x^2) + I(v_CA21_452^2) + I(low_income^2) +
                    food_density:pop_density.x + food_density:v_CA21_452 + food_density:low_income +
                    pop_density.x:v_CA21_452 + pop_density.x:low_income +
                    v_CA21_452:low_income, 
                  data = training_data_res)

n <- nrow(training_data)
R_squared <- summary(white_model)$r.squared
white_test_statistic <- n * R_squared

p_value <- pchisq(white_test_statistic, df = length(coef(white_model)) - 1, lower.tail = FALSE)

cat("White's Test Statistic:", white_test_statistic, "\n")
cat("p-value:", p_value, "\n")
```

```{r}
bp_model <- lm(residuals_squared ~ 
                 food_density + pop_density.x + v_CA21_452 + low_income, 
               data = training_data_res, na.action = na.omit)

R_squared_bp <- summary(bp_model)$r.squared
bp_test_statistic <- n * R_squared_bp

p_value_bp <- pchisq(bp_test_statistic, df = length(coef(bp_model)) - 1, lower.tail = FALSE)

cat("Breusch-Pagan Test Statistic:", bp_test_statistic, "\n")
cat("Breusch-Pagan Test p-value:", p_value_bp, "\n")
```

```{r, results='asis'}
# Calculate robust standard errors and re-run the model with robust covariance matrix
robust_se <- coeftest(reg_all_vars, vcov = vcovHC(reg_all_vars, type = "HC"))

stargazer(robust_se, type = "latex", 
          covariate.labels = c("Food Program Density", "Population Density", "Household Size", "Low Income", "Food Density:Low Income"),
          dep.var.labels = c("Crime Density")
         )
```

Gotta make new table of results with robust se for table later, keep other specs same.

```{r}
stargazer(robust_se, reg_no_interaction, reg_food_crime_low_income, type = "latex", 
          covariate.labels = c("Food Program Density", "Population Density", "Household Size", "Low Income", "Food Density:Low Income"),
          dep.var.labels = c("Crime Density")
         )
```

## Summary statistics

data_for_model \<- st_set_geometry(training_data, NULL)

model_vars \<- all.vars(formula(reg_all_vars))

subset_data \<- data_for_model %\>% select(all_of(model_vars))

readable_names \<- c(

"food_density" = "Food Program Density",

"pop_density.x" = "Population Density",

"v_CA21_452" = "Average Household Size",

"low_income" = "Low Income",

"crime_density" = "Crime Density"

)

subset_data \<- subset_data %\>% rename(any_of(readable_names))

summary_statistics \<- subset_data %\>%

summarise(across(everything(), list(

mean = \~mean(. , na.rm = TRUE),

sd = \~sd(. , na.rm = TRUE),

min = \~min(. , na.rm = TRUE),

max = \~max(. , na.rm = TRUE),

median = \~median(. , na.rm = TRUE)

))) %\>%

pivot_longer(cols = everything(), names_to = "variable_stat", values_to = "value") %\>%

separate(variable_stat, into = c("variable", "stat"), sep = "\_", extra = "drop") %\>%

pivot_wider(names_from = stat, values_from = value, values_fn = list(value = \~first(.)))

```{r}
summary_data <- st_set_geometry(training_data, NULL) %>%
  select(
    food_density, 
    low_income, 
    pop_density.x, 
    v_CA21_452
  )


# Create summary table
summary_table <- data.frame(
  Mean = sapply(summary_data, mean, na.rm = TRUE),
  SD = sapply(summary_data, sd, na.rm = TRUE),
  Max = sapply(summary_data, max, na.rm = TRUE), 
  Median = sapply(summary_data, median, na.rm = TRUE)
)

summary_table <- round(summary_table, 4)


print(summary_table)
```
