---
output:
  pdf_document: 
    latex_engine: lualatex
  html_document: default
header-includes:
  - \usepackage{graphicx}
---

```{r setup, include=FALSE}
lapply(c("tidyverse", "cancensus", "knitr", "readr", "sf", 
         "geojsonsf", "paletteer", "kableExtra", 
         "broom", "here", "stargazer"), library, character.only = TRUE)
```

## Census data

We load previously loaded census data. The code for fetching this data is also shown in this section.

```{r, eval = FALSE}
#| execute: false
load(here("API_KEY.rda"))

options(cancensus.api_key = api_key)
options(cancensus.cache_path = "cache")

vectors <- c("v_CA21_1", "v_CA21_6", "v_CA21_452", "v_CA21_449", "v_CA21_1040", "v_CA21_1085", "v_CA21_905")

region_DA <- c("59154012", "59154105", "59154090", "59150936", "59154101", "59154104", 
               "59154035", "59154103", "59154102", "59154034", "59150945", "59154091", 
               "59154093", "59154099", "59150946", "59154100", "59154078", "59154079", 
               "59154082", "59154081", "59154080", "59150939", "59150938", "59154083", 
               "59154095", "59154084", "59150941", "59150942", "59154085", "59154088", 
               "59154087", "59154089", "59154097", "59154098", "59154096", "59154092", 
               "59154013", "59150952")

census_data <- get_census(
  dataset = "CA21",
  regions = list(CSD = "5915022", DA = region_DA),
  vectors = vectors,
  labels = "detailed",
  geo_format = "sf",
  level = "DA"
)

census_data <- census_data %>%
  mutate(pop_density = `v_CA21_1: Population, 2021` / `Shape Area`)

can_api_key <- ""
save(census_data, file = "../../data/census.rda")
```

```{r table_1}
load(here("data/census.rda"))

n <- nrow(census_data)
kable(head(census_data[, 1:5]), format = "latex", booktabs = TRUE, caption = "Census Data, 2021 (truncated rows and columns)") %>%
  kable_styling(latex_options = c("striped", "hold_position"))
```

## Food data

```{r}
food_data <- st_read(here("data/free-and-low-cost-food-programs.shp"))  %>%
  select(
    "program_nam",
    "program_sta",
    "meal_cost",
    "local_areas",
    "latitude",
    "longitude",
    "geometry"
  ) %>%
  drop_na("latitude", "longitude") %>%
  # set to wgs 84 as per can census
  st_set_crs(4326)
```

```{r table_2}
# Food data processing
food_count <- food_data %>%
  st_set_geometry(NULL) %>% 
  group_by(local_areas) %>%
  summarise(count = n(), .groups = "drop")

food_data_count <- food_data %>%
  left_join(food_count, by = "local_areas") %>%
  distinct(local_areas, .keep_all = TRUE) # one row per neighbourhood

combo_food_census <- census_data %>%
  st_join(food_data_count)

census_data_food <- combo_food_census %>%
  mutate(program_count = replace_na(count, 0),
         food_density = program_count / `Shape Area`)

head(census_data_food[, 1:5]) %>%
  kable(format = "latex", booktabs = TRUE, caption = "Food Data merged with Census Data, 2021 (truncated rows and columns)") %>%
  kable_styling(latex_options = c("striped", "hold_position"))
```

\_

## Crime data

```{r}
crime <- read_csv(here("data/crime_data_all_neighborhoods.csv"), show_col_types = FALSE) %>%
  mutate(TYPE = as_factor(TYPE),
         HUNDRED_BLOCK = as_factor(HUNDRED_BLOCK),
         NEIGHBOURHOOD = as_factor(NEIGHBOURHOOD)) %>%
  filter(!is.na(X) & !is.na(Y))

crime_data <- st_as_sf(crime, coords = c("X", "Y"), crs = "+proj=utm +zone=10") %>%
  st_transform(crs = "+proj=longlat +datum=WGS84")
```

```{r}
intersections <- st_is_within_distance(census_data, crime_data, sparse = FALSE, dist = 5)

crimes_contained <- rowSums(intersections, dims = 1)

census_data_crime <- census_data %>%
  cbind(crimes_contained) %>%
  mutate(crime_density = crimes_contained / Shape.Area)

unique_crimes <- unique(crime_data$TYPE)

for (type in unique_crimes) {
  type_data <- crime_data %>% filter(TYPE == type)
  intersections <- st_is_within_distance(census_data, type_data, sparse = FALSE, dist = 5)
  sum <- rowSums(intersections, dims = 1)
  df <- as.data.frame(sum)
  census_data_crime <- census_data_crime %>% cbind(df$sum) %>% rename_with(~ paste0("crimes_", type), df.sum)
}
```

## Training data

Now we merge all of our data-sets into one table so we can feed it into our model.

```{r}
training_data <- st_join(census_data_crime, census_data_food)

training_data <- training_data %>%
  rename_with(
    ~ gsub(":.*$", "", .),  # Remove everything after the colon, including the colon
    starts_with("v_CA21")  # Apply only to columns starting with "v_CA21"
  )

median_lico_at <- median(training_data$v_CA21_1085, na.rm = TRUE)

training_data <- training_data %>%
  mutate(
    low_income = ifelse(
      v_CA21_1085 > median_lico_at, 
      1, 
      0))


training_data[is.na(training_data)] <- 0
```

# Model

## Specification

```{r}
# Specification models
# 1. Model with all variables (including interaction term)
reg_all_vars <- lm(crime_density ~ 
                   food_density:low_income + 
                   food_density + 
                   pop_density.x + 
                   v_CA21_452 +
                   low_income, 
                 data = st_set_geometry(training_data, NULL))

# 2. Model without the interaction term
reg_no_interaction <- lm(crime_density ~ 
                         food_density + 
                         pop_density.x + 
                         v_CA21_452 +
                         low_income, 
                       data = st_set_geometry(training_data, NULL))

# 3. Model with only food density, crime density, and low income
reg_food_crime_low_income <- lm(crime_density ~ 
                                food_density + 
                                low_income, 
                              data = st_set_geometry(training_data, NULL))

# 4. Model with all variables (including interaction term) but replacing v_CA21_452 with v_CA21_449
reg_all_vars_449 <- lm(crime_density ~ 
                       food_density:low_income + 
                       food_density + 
                       pop_density.x + 
                       v_CA21_449 +  # Replaced v_CA21_452 with v_CA21_449
                       low_income, 
                     data = st_set_geometry(training_data, NULL))

models_spec <- list(
  "All Variables" = reg_all_vars,
  "Without Interaction" = reg_no_interaction,
  "Food Density, Crime, Low Income" = reg_food_crime_low_income
)

# Summary for each specification tested
summary_all_vars <- summary(reg_all_vars)
summary_no_interaction <- summary(reg_no_interaction)
summary_food_crime_low_income <- summary(reg_food_crime_low_income)
summary_all_vars_449 <- summary(reg_all_vars_449)

# Add to the list of model summaries
model_summaries <- list(
  "All Variables" = summary_all_vars,
  "Without Interaction" = summary_no_interaction,
  "Food Density, Crime, Low Income" = summary_food_crime_low_income,
  "All Variables (with v_CA21_449)" = summary_all_vars_449
)
```

# RESULTS FROM SPEC

### Function definitions

We needed to manipulate our regression model data into formatted tables that emulated those from the package `stargazer` as we were having significant alignment issues between the data and rendered tables.

```{r}
all_variables <- unique(c(
  names(coef(reg_all_vars)),
  names(coef(reg_no_interaction)),
  names(coef(reg_food_crime_low_income))
))

all_variables_escaped <- gsub("_", "\\\\_", all_variables)  # Escape underscores
all_variables_escaped <- paste0("\\texttt{", all_variables_escaped, "}")  # Wrap in \texttt{} for LaTeX

add_stars <- function(estimates, ses, p_values) {
  significance_levels <- ifelse(p_values < 0.01, "***",
                          ifelse(p_values < 0.05, "**",
                          ifelse(p_values < 0.1, "*", "")))
  formatted <- sprintf("%.3f (%.3f)%s", estimates, ses, significance_levels)
  return(formatted)
}

extract_summary <- function(model, all_vars) {
  coefs <- coef(model)
  ses <- sqrt(diag(vcov(model)))
  p_values <- coef(summary(model))[, 4]  # Extract p-values from the summary

  # Create placeholders for all variables
  aligned_coefs <- setNames(rep(NA, length(all_vars)), all_vars)
  aligned_ses <- setNames(rep(NA, length(all_vars)), all_vars)
  aligned_p_values <- setNames(rep(NA, length(all_vars)), all_vars)

  # Fill with existing coefficients, standard errors, and p-values
  for (var in names(coefs)) {
    aligned_coefs[var] <- coefs[var]
    aligned_ses[var] <- ses[var]
    aligned_p_values[var] <- p_values[var]
  }

  # Replace missing coefficients with 0
  aligned_coefs[is.na(aligned_coefs)] <- 0

  # Add stars to coefficients based on significance levels
  significance_levels <- ifelse(aligned_p_values < 0.01, "***",
                          ifelse(aligned_p_values < 0.05, "**",
                          ifelse(aligned_p_values < 0.1, "*", "")))

  # Format coefficients and standard errors for display
  formatted <- ifelse(is.na(aligned_ses),
                      sprintf("%.3f", aligned_coefs),
                      sprintf("%.3f (%.3f)%s", aligned_coefs, aligned_ses, significance_levels))

  # Replace missing standard errors with blank strings
  formatted[is.na(aligned_ses)] <- " "

  # Escape underscores for LaTeX compatibility and wrap in \texttt{} for table display
  formatted <- gsub("_", "\\\\_", formatted)
  formatted <- paste0("\\texttt{", formatted, "}")

  return(formatted)
}

extract_model_info <- function(model, covariate_labels, model_title) {
  coefs <- coef(model)
  ses <- sqrt(diag(vcov(model)))
  p_values <- coef(summary(model))[, 4]  # Extract p-values from the summary
  
  # Format coefficients, standard errors, and significance stars
  significance_levels <- ifelse(p_values < 0.01, "***",
                                ifelse(p_values < 0.05, "**",
                                       ifelse(p_values < 0.1, "*", "")))
  
  formatted <- sprintf("%.3f (%.3f)%s", coefs, ses, significance_levels)
  
  # Combine covariate labels and the formatted coefficients
  result <- data.frame(
    Variable = covariate_labels,
    Estimate = formatted,
    stringsAsFactors = FALSE
  )
  
  # Add model title
  result <- rbind(data.frame(Variable = model_title, Estimate = "", stringsAsFactors = FALSE), result)
  
  return(result)
}



# Extract R-squared, F-statistics, and number of observations from each model
r_squared <- c(
  summary(reg_all_vars)$r.squared,
  summary(reg_no_interaction)$r.squared,
  summary(reg_food_crime_low_income)$r.squared
)

f_statistic <- c(
  summary(reg_all_vars)$fstatistic[1],
  summary(reg_no_interaction)$fstatistic[1],
  summary(reg_food_crime_low_income)$fstatistic[1]
)

n_obs <- c(
  length(reg_all_vars$fitted.values),
  length(reg_no_interaction$fitted.values),
  length(reg_food_crime_low_income$fitted.values)
)


model1_aligned <- extract_summary(reg_all_vars, all_variables)
model2_aligned <- extract_summary(reg_no_interaction, all_variables)
model3_aligned <- extract_summary(reg_food_crime_low_income, all_variables)
```

### Data manipulation

```{r}
aligned_table <- data.frame(
  Variable = all_variables_escaped,
  `Model 1` = model1_aligned,
  `Model 2` = model2_aligned,
  `Model 3` = model3_aligned,
  stringsAsFactors = FALSE
)

print(aligned_table)

summary_table <- data.frame(
  Variable = c("R squared", "F statistic", "Observations"),
  `Model 1` = c(sprintf("%.3f", r_squared[1]), sprintf("%.1f", f_statistic[1]), n_obs[1]),
  `Model 2` = c(sprintf("%.3f", r_squared[2]), sprintf("%.1f", f_statistic[2]), n_obs[2]),
  `Model 3` = c(sprintf("%.3f", r_squared[3]), sprintf("%.1f", f_statistic[3]), n_obs[3]),
  stringsAsFactors = FALSE
)

# Combine both tables
final_table <- rbind(aligned_table, summary_table)
print(final_table)
```

```{r}
covariate_labels <- c(
  "(Intercept)", 
  "Food Program Density: Low Income",
  "Food Program Density",
  "Population Density", 
  "Average Household Size (v\\_CA21\\_452)",
  "Low Income"
)
covariate_labels_2 <- c(
  "(Intercept)",
  "Food Program Density",
  "Population Density", 
  "Average Household Size (v\\_CA21\\_452)",
  "Low Income"
)

covariate_labels_3 <- c(
  "(Intercept)",
  "Food Program Density",
  "Low Income"
)

model_1_table <- extract_model_info(
  model = reg_all_vars,
  covariate_labels = covariate_labels,
  model_title = "Spec 1: All Variables"
)

model_2_table <- extract_model_info(
  model = reg_no_interaction,
  covariate_labels = covariate_labels_2,
  model_title = "Spec 2: Omitted Interaction Term"
)

model_3_table <- extract_model_info(
  model = reg_food_crime_low_income,
  covariate_labels = covariate_labels_3,
  model_title = "Spec 3: Food Density and Low Income"
)
```

### Regression table

```{r}
# Render the final table in a stargazer-like format using knitr::kable
kable(
  final_table,
  format = "latex",
  col.names = c("Variable", "Model 1", "Model 2", "Model 3"),
  caption = "Regression Specifications",
  align = "lccc",
  booktabs = TRUE,
  escape = FALSE
) %>%
  kable_styling(latex_options = c("hold_position", "striped")) %>%
  add_header_above(c(" " = 1, "Dependent Variable: Crime Density" = 3)) %>%
  footnote(general = "* p < 0.1; ** p < 0.05; *** p < 0.01", 
           general_title = "Note:", 
           footnote_as_chunk = TRUE, 
           escape = FALSE)
```

### Model 1:

```{r, results='asis'}
# Model 1: reg_all_vars
kable(
  model_1_table,
  format = "latex",
  col.names = c("Variable", "Estimate"),
  caption = "All Variables (Model 1)",
  align = "lc",
  booktabs = TRUE,
  escape = FALSE
) %>%
  kable_styling(latex_options = c("hold_position", "striped")) %>%
  footnote(general = "* p < 0.1; ** p < 0.05; *** p < 0.01", 
           general_title = "Note:", 
           footnote_as_chunk = TRUE, 
           escape = FALSE)
```

### Model 2:

```{r, results='asis'}
# Model 2: reg_no_interaction
kable(
  model_2_table,
  format = "latex",
  col.names = c("Variable", "Estimate"),
  caption = "Omitted Interaction Term (Model 2)",
  align = "lc",
  booktabs = TRUE,
  escape = FALSE
) %>%
  kable_styling(latex_options = c("hold_position", "striped")) %>%
  footnote(general = "* p < 0.1; ** p < 0.05; *** p < 0.01", 
           general_title = "Note:", 
           footnote_as_chunk = TRUE, 
           escape = FALSE)
```

### Model 3:

```{r, results='asis'}
# Model 3: reg_food_crime_low_income
kable(
  model_3_table,
  format = "latex",
  col.names = c("Variable", "Estimate"),
  caption = "Food Density and Low Income (Model 3)",
  align = "lc",
  booktabs = TRUE,
  escape = FALSE
) %>%
  kable_styling(latex_options = c("hold_position", "striped")) %>%
  footnote(general = "* p < 0.1; ** p < 0.05; *** p < 0.01", 
           general_title = "Note:", 
           footnote_as_chunk = TRUE, 
           escape = FALSE)
```

## Robustness

We split the data into sub-groups based on

```{r}
census_data_centroids <- census_data %>%
  mutate(
    centroid = st_centroid(geometry),
    latitude = st_coordinates(centroid)[, 2],
    longitude = st_coordinates(centroid)[, 1]
  ) %>%
  st_set_geometry(NULL)  # Remove geometry for k-means input

# Normalize latitude and longitude for clustering
census_data_normalized <- census_data_centroids %>%
  mutate(
    latitude_scaled = scale(latitude),
    longitude_scaled = scale(longitude)
  )


set.seed(123) # For reproducibility
k <- 4

kmeans_result <- kmeans(census_data_normalized[, c("latitude_scaled", "longitude_scaled")], centers = k)

# Add cluster assignments to the original data
census_data$cluster <- kmeans_result$cluster

# Now, merge the cluster assignments into your training data
training_data <- st_join(training_data, 
                         census_data %>% select(cluster, geometry), 
                         left = TRUE)
```

```{r}
# Run the model for each cluster
for (i in 1:k) {
  subset_data <- training_data %>% filter(cluster == i)
  
  model <- lm(crime_density ~ 
                food_density:low_income + 
                food_density + 
                pop_density.x + 
                v_CA21_452 +
                low_income, 
              data = st_set_geometry(subset_data, NULL))
  
  cat(paste("\nModel Summary for Cluster", i, ":\n"))
  print(summary(model)) # Check the results for each cluster
}
```

```{r}
# Step 1: Prepare a container for the model results
cluster_summary_table <- do.call(rbind, lapply(1:k, function(i) {
  subset_data <- training_data %>% filter(cluster == i)
  
  # Run the linear model for each cluster
  model <- lm(crime_density ~ 
                food_density:low_income + 
                food_density + 
                pop_density.x + 
                v_CA21_452 +
                low_income, 
              data = st_set_geometry(subset_data, NULL))
  
  model_summary <- summary(model)
  
  # Extract coefficients, standard errors, p-values, and significance levels
  coefs <- coef(model)
  ses <- sqrt(diag(vcov(model)))
  p_values <- coef(summary(model))[, 4]
  
  # Calculate significance levels for coefficients
  significance_levels <- ifelse(p_values < 0.01, "***",
                          ifelse(p_values < 0.05, "**",
                          ifelse(p_values < 0.1, "*", "")))
  
  # Format coefficients with standard errors and significance levels
  estimates <- sprintf("%.3f (%.3f)%s", coefs, ses, significance_levels)

  # Create a data frame for the cluster summary
  data.frame(
    Cluster = paste("Cluster", i),
    `(Intercept)` = estimates[1],
    `food_density` = ifelse(length(estimates) >= 2, estimates[2], NA),
    `pop_density.x` = ifelse(length(estimates) >= 3, estimates[3], NA),
    `v_CA21_452` = ifelse(length(estimates) >= 4, estimates[4], NA),
    `low_income` = ifelse(length(estimates) >= 5, estimates[5], NA),
    `food_density:low_income` = ifelse(length(estimates) >= 6, estimates[6], NA),
    R_squared = sprintf("%.3f", model_summary$r.squared),
    F_statistic = ifelse(!is.null(model_summary$fstatistic), sprintf("%.1f", model_summary$fstatistic[1]), NA),
    Observations = length(model$fitted.values),
    stringsAsFactors = FALSE
  )
}))
```

```{r, results='asis'}
kable(
  cluster_summary_table,
  format = "latex",
  col.names = c("Cluster", "(Intercept)", "Food Density", "Population Density", 
                "Average Household Size (v\\_CA21\\_452)", "Low Income", "Food Density: Low Income",
                "R Squared", "F Statistic", "Observations"),
  caption = "Robustness Check: Model Results for Each Cluster",
  align = "lcccccccc",
  booktabs = TRUE,
  escape = FALSE
) %>%
  kable_styling(latex_options = c("hold_position", "striped")) %>%
  add_header_above(c(" " = 1, "Cluster-wise Model Summary" = 9)) %>%
  footnote(general = "* p < 0.1; ** p < 0.05; *** p < 0.01", 
           general_title = "Note:", 
           footnote_as_chunk = TRUE, 
           escape = FALSE)
```
