---
title: "Final Proposal (Group 14) - Advanced Track"
author: "Anishka Fernandopulle, Radhika Iyer, Sidharth Mirchandani, Hajra Ojha"
output:
  html_document: 
    theme: cerulean
    df_print: kable
  word_document: default
  pdf_document: 
    fig_width: 5
    fig_height: 4
    fig_caption: true
---

```{r, results='hide',message=FALSE, echo=FALSE}
library(tidyverse)
library(cancensus)
library(knitr)
library(readr)
library(sf)
library(geojsonsf)
library(paletteer)
library(kableExtra)

options(knitr.kable.NA = 'NA')

load("API_KEY.rda")

options(cancensus.api_key = api_key)
options(cancensus.cache_path = "cache")

vectors <- c("v_CA21_1",
             "v_CA21_6",
             "v_CA21_449",
             "v_CA21_1040",
             "v_CA21_1085",
             "v_CA21_905")

census_data <- get_census(
  dataset = "CA21",
  regions = list(
    CSD = "5915022",
    DA = c(
      "59154012",
      "59154105",
      "59154090",
      "59150936",
      "59154101",
      "59154104",
      "59154035",
      "59154103",
      "59154102",
      "59154034",
      "59150945",
      "59154091",
      "59154093",
      "59154099",
      "59150946",
      "59154100",
      "59154078",
      "59154079",
      "59154082",
      "59154081",
      "59154080",
      "59150939",
      "59150938",
      "59154083",
      "59154095",
      "59154084",
      "59150941",
      "59150942",
      "59154085",
      "59154088",
      "59154087",
      "59154089",
      "59154097",
      "59154098",
      "59154096",
      "59154092",
      "59154013",
      "59150952"
    )
  ),
  vectors = vectors,
  labels = "detailed",
  geo_format = "sf",
  level = "DA"
)

can_api_key <- ""
save(census_data, file = "data/census.rda")

food_data <- st_read("data/free-and-low-cost-food-programs.shp")  %>%
  select(
    "program_nam",
    "program_sta",
    "meal_cost",
    "local_areas",
    "latitude",
    "longitude",
    "geometry"
  ) %>%
  drop_na("latitude", "longitude") %>%
  # set to wgs 84 as per can census
  st_set_crs(4326)

head(food_data)
head(food_data$geometry)

crime <- read_csv("data/crime_data_all_neighborhoods.csv")
colnames(crime) <- tolower(make.names(colnames(crime)))

head(crime)
# now we mutate crime data to add shapefile information
# first we create as a utm shapefile due to the data structure,
# then convert to lat/long like census_data and food_data
# crs = 32610 : utm 10
# crs = 4326 : lat/long WGS84
crime_data <- crime %>%
  st_as_sf(coords = c("x", "y"), crs = 32610) %>%
  st_transform(crs = 4326)

# we need to remove (0,0) values for plotting as they cause errors on the axis
# to do that we extract coords then filter where the Y component == 0
crime_coords <- st_coordinates(crime_data)
crime_data <- crime_data[crime_coords[, 2] != 0, ]

census_geom <- st_geometry(census_data)
census_centroids <- st_centroid(census_geom)

distance_food <- st_distance(census_centroids, food_data)

intersections_food <- st_intersects(census_data, food_data, sparse = FALSE)
intersections_food_500 <- st_is_within_distance(census_data, food_data, sparse = FALSE, dist = 500)
food_contained <- rowSums(intersections_food, dims = 1)
food_within_500 <- rowSums(intersections_food_500, dims = 1)




intersections_crime <- st_intersects(census_data, crime_data, sparse = FALSE)
intersections_crime_500 <- st_is_within_distance(census_data, crime_data, sparse = FALSE, dist = 500)
crime_contained <- rowSums(intersections_crime, dims = 1)
crime_within_500 <- rowSums(intersections_crime_500, dims = 1)


vectors <- c("v_CA21_1", "v_CA21_6", "v_CA21_449", "v_CA21_1040", "v_CA21_1085", "v_CA21_905")

summary_statistics_census <- list()
combo_statistics_census <- data.frame()

# create summary table for every vector in census data
# sources for manipulating through the list:
# https://stackoverflow.com/a/25839767
# https://stackoverflow.com/a/75671893

for (vector in vectors) {
  summary_table <- census_data %>%
    st_drop_geometry() %>%  # Removes geometry column
    select(matches(vector)) %>%
    pivot_longer(cols = everything(), names_to = "variable", values_to = "value") %>%
    filter(!is.na(value)) %>%
    summarize(
      mean = mean(value, na.rm = TRUE),
      sd = sd(value, na.rm = TRUE),
      max = max(value, na.rm = TRUE),
      min = min(value, na.rm = TRUE),
      .groups = "drop"
    ) %>%
    mutate(variable = vector) %>%
    select(variable, everything())
  
  summary_statistics_census[[vector]] <- summary_table
  
  combo_statistics_census <- rbind(combo_statistics_census, summary_table)
}

food_count <- food_data %>%
  st_set_geometry(NULL) %>% 
  group_by(local_areas) %>%
  summarise(count = n(), .groups = "drop")

food_data_count <- food_data %>%
  left_join(food_count, by = "local_areas") %>%
  distinct(local_areas, .keep_all = TRUE) # one row per neighbourhood

combo_food_census <- census_data %>%
  st_join(food_data_count)

combo_food_census <- combo_food_census %>%
 mutate(program_count = replace_na(count, 0),
        area = st_area(geometry),
        density = program_count / (area * 1e-6))

summary_statistics_food <- combo_food_census %>%
  select(local_areas, program_count, density) %>%
  group_by(local_areas) %>%
  summarise(
    count = sum(program_count, na.rm = TRUE),
    mean = mean(density, na.rm = TRUE),
    sd = sd(density, na.rm = TRUE),
    max = max(density, na.rm = TRUE),
    min = min(density, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  select(-geometry, -count, local_areas, mean, sd, max, min)%>%
  st_set_geometry(NULL)

combo_crime_census <- st_join(census_data, crime_data)

summary_statistics_crime <- combo_crime_census %>%
  group_by(neighbourhood) %>%
  reframe(
    count = n(),
    area = st_area(geometry),
    crime_density = count / (area * 1e-6)  # Convert area to square kilometers
  ) %>%
  group_by(neighbourhood) %>%
  summarise(
    total_crimes = first(count),
    mean = mean(crime_density, na.rm = TRUE),
    sd = sd(crime_density, na.rm = TRUE),
    max = max(crime_density, na.rm = TRUE),
    min = min(crime_density, na.rm = TRUE)
  ) %>%
  ungroup() %>%
  select(neighbourhood, mean, sd, max, min)
```

```{r}
library(tidyverse)
library(cancensus)
library(knitr)
library(readr)
library(sf)
library(geojsonsf)
library(paletteer)
library(units)
library(broom)
library(kableExtra)
```

### Loading census using API key

#### First we load our api key and create our vector of columns we would like to extract

```{r}
load("API_KEY.rda")

options(cancensus.api_key = api_key)
options(cancensus.cache_path = "cache")

vectors <- c("v_CA21_1",
             "v_CA21_6",
             "v_CA21_449",
             "v_CA21_1040",
             "v_CA21_1085",
             "v_CA21_905")
```

#### Get the dataset and save to file. Clear api key out of global env. THIS DATA CONTAINS SF INFO FOR CENSUS!!!!

```{r}
census_data <- get_census(
  dataset = "CA21",
  regions = list(
    CSD = "5915022",
    DA = c(
      "59154012",
      "59154105",
      "59154090",
      "59150936",
      "59154101",
      "59154104",
      "59154035",
      "59154103",
      "59154102",
      "59154034",
      "59150945",
      "59154091",
      "59154093",
      "59154099",
      "59150946",
      "59154100",
      "59154078",
      "59154079",
      "59154082",
      "59154081",
      "59154080",
      "59150939",
      "59150938",
      "59154083",
      "59154095",
      "59154084",
      "59150941",
      "59150942",
      "59154085",
      "59154088",
      "59154087",
      "59154089",
      "59154097",
      "59154098",
      "59154096",
      "59154092",
      "59154013",
      "59150952"
    )
  ),
  vectors = vectors,
  labels = "detailed",
  geo_format = "sf",
  level = "DA"
)

can_api_key <- ""
save(census_data, file = "data/census.rda")
```

```{r}
head(census_data)
```

### (note)Then we have to load other data, transform , add sf info, then merge into dataset and add comparison functions

### Loading other datasets , renaming and standardizing columns

```{r}
food_data <- st_read("data/free-and-low-cost-food-programs.shp")  %>%
  select(
    "program_nam",
    "program_sta",
    "meal_cost",
    "local_areas",
    "latitude",
    "longitude",
    "geometry"
  ) %>%
  drop_na("latitude", "longitude") %>%
  # set to wgs 84 as per can census
  st_set_crs(4326)

head(food_data)
head(food_data$geometry)
```

```{r}
crime <- read_csv("data/crime_data_all_neighborhoods.csv")
colnames(crime) <- tolower(make.names(colnames(crime)))

head(crime)
# now we mutate crime data to add shapefile information
# first we create as a utm shapefile due to the data structure,
# then convert to lat/long like census_data and food_data
# crs = 32610 : utm 10
# crs = 4326 : lat/long WGS84
crime_data <- crime %>%
  st_as_sf(coords = c("x", "y"), crs = 32610) %>%
  st_transform(crs = 4326)

# we need to remove (0,0) values for plotting as they cause errors on the axis
# to do that we extract coords then filter where the Y component == 0
crime_coords <- st_coordinates(crime_data)
crime_data <- crime_data[crime_coords[, 2] != 0, ]

head(crime_data)
head(st_geometry(crime_data))
```

### combined table

```{r}
# need to add density from summary statistics crime for neighbourhood
combined_table <- st_join(combo_food_census, combo_crime_census, join = st_intersects)

combined_table <- combined_table %>%
  select(-ends_with(".y")) %>%
  rename_with(~ gsub("\\.x$", "", .), ends_with(".x"))

combined_table <- combined_table %>%
  left_join(summary_statistics_crime %>% select(neighbourhood, mean), 
            by = "neighbourhood") %>%
  rename(crime_density = mean) %>%
  rename(food_density = density)

combined_table <- combined_table %>%
  rename_with(
    ~ gsub(":.*$", "", .),  # Remove everything after the colon, including the colon
    starts_with("v_CA21")  # Apply only to columns starting with "v_CA21"
  )

columns_to_drop_units <- c(
  "crime_density",
  "food_density",
  "v_CA21_1040",
  "v_CA21_1",
  "v_CA21_449"
)

combined_table <- combined_table %>%
  mutate(across(all_of(columns_to_drop_units), function(col) {
    if (inherits(col, "units")) drop_units(col) else col
  }))
# View the result
head(combined_table)
```

### dummy variable low income

```{r}
# calculate the median LICO-AT across all neighborhoods
median_lico_at <- median(combined_table$v_CA21_1085, na.rm = TRUE)

# Add the low_income column
combined_table <- combined_table %>%
  mutate(
    low_income = ifelse(
      v_CA21_1085 > median_lico_at, 
      1, 
      0
    )
  )

# View the updated table
head(combined_table)


```

### Shapefile distance calculations

```{r}
# example statistics from jonathon
census_geom <- st_geometry(census_data)
census_centroids <- st_centroid(census_geom)

distance_food <- st_distance(census_centroids, food_data)

intersections_food <- st_intersects(census_data, food_data, sparse = FALSE)
intersections_food_500 <- st_is_within_distance(census_data, food_data, sparse = FALSE, dist = 500)
food_contained <- rowSums(intersections_food, dims = 1)
food_within_500 <- rowSums(intersections_food_500, dims = 1)

mean(food_contained)
mean(food_within_500)
```

#### this cell takes a long time to run, idk why right now so avoid running it, all other calcs work independent (it does finish running though)

```{r}
distance_crime <- st_distance(census_centroids, crime_data)
```

```{r}
intersections_crime <- st_intersects(census_data, crime_data, sparse = FALSE)
intersections_crime_500 <- st_is_within_distance(census_data, crime_data, sparse = FALSE, dist = 500)
crime_contained <- rowSums(intersections_crime, dims = 1)
crime_within_500 <- rowSums(intersections_crime_500, dims = 1)


mean(crime_contained)
mean(crime_within_500)
```

### Visualizing our data

```{r}
can_map <- ggplot() +
  geom_sf(data = census_data, color = "blue") +
  theme_minimal() +
  labs(title = "Census data mapping", )

crime_map <- ggplot() +
  geom_sf(
    data = crime_data,
    aes(color = type),
    alpha = 0.8,
    size = 0.5,
    fill = "white"
  ) +
  theme_minimal() +
  labs(title = "Crime data mapping",
       fill = "Legend",
       color = "Crime Type") +
  theme(legend.position = "right") +
  scale_color_paletteer_d("ggthemes::Classic_Purple_Gray_12")

# some extra themes
# scale_color_paletteer_d("ggprism::autumn_leaves")
# scale_color_paletteer_d("beyonce::X70")
# scale_color_paletteer_d("impressionist.colors::te_aa_no_areois")
# scale_color_paletteer_d("ggthemes::Classic_Purple_Gray_12")

food_map <- ggplot() +
  geom_sf(data = food_data, color = "blue") +
  theme_minimal() +
  labs(title = "Food center mapping")

can_map
crime_map
food_map
```

```{r}
crime_food_plot <- ggplot() +
  geom_sf(data = census_data, color = "black") +
  geom_sf(
    data = crime_data,
    aes(color = "geometry"),
    size = 0.4,
    alpha = 0.2
  ) +
  geom_sf(
    data = food_data,
    aes(fill = "yellow"),
    color = "black",
    size = 2,
    alpha = 0.8,
    shape = 21
  ) +
  coord_sf() +
  labs(title = "Free and Low Cost Food Programs plotted against Crime, Vancouver") +
  theme_minimal() +
  theme(legend.position = "right") +
  scale_fill_manual(
    name = "",
    values = "yellow",
    labels = c("Food Programs")
  ) +
  scale_color_manual(name = "",
                     values = "red",
                     labels = c("Crime")) +
  # increase point size for legend only
  guides(fill = guide_legend(override.aes = list(size = 5)),
         color = guide_legend(override.aes = list(size = 5, alpha = 0.5)))

crime_food_plot
```

```{r}
lim_at_plot <- ggplot() +
  geom_sf(
    data = census_data,
    aes(fill = `v_CA21_1040: Prevalence of low income based on the Low-income measure, after tax (LIM-AT) (%)`),
    color = "black",
    size = 0.2
  ) +
  scale_fill_paletteer_c("viridis::viridis", 
                         name = "Low Income (%)",
                         limits = c(0, 100)) +
  labs(title = "Prevalence of low income based on the Low-income measure, after tax") +
  theme_minimal() +
  theme(legend.position = "right")

lim_at_plot 

lico_at_plot <- ggplot() +
  geom_sf(
    data = census_data,
    aes(fill = `v_CA21_1085: Prevalence of low income based on the Low-income cut-offs, after tax (LICO-AT) (%)`),
    color = "black",
    size = 0.2
  ) +
  scale_fill_paletteer_c("viridis::viridis", 
                         name = "Low Income (%)",
                         limits = c(0, 100)) +
  labs(title = "Prevalence of low income based on the Low-income cut-offs, after tax") +
  theme_minimal() +
  theme(legend.position = "right")

lico_at_plot
```

```{r}
lim_crime_food_plot <- ggplot() +
  # lim-at + can map from census data
  geom_sf(
    data = census_data,
    aes(fill = `v_CA21_1040: Prevalence of low income based on the Low-income measure, after tax (LIM-AT) (%)`),
    color = "black",
    size = 0.2
  ) +
  # crime mapping
  geom_sf(
    data = crime_data,
    color = "red",
    size = 0.3,
    alpha = 0.2
  ) +
  geom_sf(
    data = food_data,
    fill = "yellow",
    color = "black",
    size = 2,
    alpha = 0.8,
    shape = 21
  ) +
  # lim-at gradient
  scale_fill_paletteer_c("viridis::viridis", 
                         name = "Low Income (%)",
                         limits = c(0, 100)) +
  labs(title = "Low Income Prevalence + Crime and Food Centers (LIM-AT)") +
  theme_minimal() +
  theme(legend.position = "right")


lim_crime_food_plot                              
```

```{r}
lico_crime_food_plot <- ggplot() +
  # Low income measure (LIM-AT) as a continuous fill
  geom_sf(
    data = census_data,
    aes(fill = `v_CA21_1085: Prevalence of low income based on the Low-income cut-offs, after tax (LICO-AT) (%)`),
    color = "black",
    size = 0.2
  ) +
  # Crime data, set color to red without aes (not mapped)
  geom_sf(
    data = crime_data,
    color = "red",
    size = 0.2,
    alpha = 0.2
  ) +
  # Food programs, set fill to yellow without aes (not mapped)
  geom_sf(
    data = food_data,
    fill = "yellow",   # Fixed yellow fill
    color = "black",
    size = 2,
    alpha = 0.8,
    shape = 21
  ) +
  # Viridis scale for continuous data
  scale_fill_paletteer_c("viridis::viridis", 
                         name = "Low Income (%)",
                         limits = c(0, 100)) +
  labs(title = "Low Income Prevalence + Crime and Food Centers (LICO-AT)") +
  theme_minimal() +
  theme(legend.position = "right")

#TODO: legends for food center


lico_crime_food_plot  
```

## Model

$$
(\text{Crime Density})_i = \beta_0+\beta_1(\text{Food Program Density}\cdot\text{Low Income})_i+\beta_2(\text{Food Program Density})_i+ \\ \beta_3(\text{Population Density})_i + \beta_4(\text{Private Households})_i + \beta_5(\text{Low Income})_i + \epsilon_i
$$

```{r}
reg_limat <- lm(crime_density ~ 
                  food_density:v_CA21_1040 + 
                  food_density + 
                  v_CA21_1 + 
                  v_CA21_449 +
                  v_CA21_1040, 
                data = combined_table)

limat_summary <- summary(reg_limat)
```

```{r}
reg_licoat <- lm(crime_density ~ 
                  food_density:v_CA21_1085 + 
                  food_density + 
                  v_CA21_1 + 
                  v_CA21_449 +
                  v_CA21_1085, 
                data = combined_table)

licoat_summary <- summary(reg_licoat)
```

```{r}
limat_coeff_table <- as.data.frame(limat_summary$coefficients)
limat_coeff_table <- cbind(Variable = rownames(limat_coeff_table), limat_coeff_table) 
rownames(limat_coeff_table) <- NULL

# Format and display the table with kable
kable(limat_coeff_table, caption = "Regression Coefficients: limat", digits = 10)
```

```{r}
licoat_coeff_table <- as.data.frame(licoat_summary$coefficients)
licoat_coeff_table <- cbind(Variable = rownames(licoat_coeff_table), licoat_coeff_table) 
rownames(licoat_coeff_table) <- NULL

# Format and display the table with kable
kable(licoat_coeff_table, caption = "Regression Coefficients: licoat", digits = 10)
```

```{r}
variables <- c(
  "`food_density`",
  "v_CA21_1",
  "v_CA21_449",
  "v_CA21_1085",
  "food_density:v_CA21_1085"
)

results <- expand_grid(Row = variables, Col = variables) %>%
  mutate(
    Formula = ifelse(
      Row == Col, NA,  # Avoid self-interaction
      paste("crime_density ~", Row, "+", Col)
    )
  ) %>%
  rowwise() %>%
  mutate(
    R_Squared = if (!is.na(Formula)) {
      mod <- lm(as.formula(Formula), data = combined_table)
      summary(mod)$r.squared
    } else {
      NA
    }
  ) %>%
  select(Row, Col, R_Squared)

# Pivot for matrix structure
results_matrix1 <- results %>%
  pivot_wider(names_from = Col, values_from = R_Squared)

# Display the table
kable(results_matrix1, 
      caption = "Regression R^2 Matrix Including Interaction Term",
      digits = 3) %>%
  kable_styling(full_width = FALSE, bootstrap_options = c("striped", "hover", "condensed"))
```

```{r}
variables <- c(
  "food_density",
  "v_CA21_1",
  "v_CA21_449",
  "v_CA21_1040",
  "food_density:v_CA21_1040"
  )

# Generate pairwise regression results including interaction
results <- expand_grid(Row = variables, Col = variables) %>%
  mutate(
    Formula = ifelse(
      Row == Col, NA,  # Avoid self-interaction
      paste("crime_density ~", Row, "+", Col)
    )
  ) %>%
  rowwise() %>%
  mutate(
    R_Squared = if (!is.na(Formula)) {
      mod <- lm(as.formula(Formula), data = combined_table)
      summary(mod)$r.squared
    } else {
      NA
    }
  ) %>%
  select(Row, Col, R_Squared)

# Pivot for matrix structure
results_matrix2 <- results %>%
  pivot_wider(names_from = Col, values_from = R_Squared)

# Display the table
kable(results_matrix2, 
      caption = "Regression R^2 Matrix Including Interaction Term",
      digits = 3) %>%
  kable_styling(full_width = FALSE, bootstrap_options = c("striped", "hover", "condensed"))
```

```{r}
reg_dummy <- lm(crime_density ~ 
                  food_density:low_income + 
                  food_density + 
                  v_CA21_1 + 
                  v_CA21_449 +
                  low_income, 
                data = combined_table)

reg_dummy_summary <- summary(reg_dummy)

dummy_coeff_table <- as.data.frame(reg_dummy_summary$coefficients)
dummy_coeff_table <- cbind(Variable = rownames(dummy_coeff_table), dummy_coeff_table) 
rownames(dummy_coeff_table) <- NULL

# Format and display the table with kable
kable(dummy_coeff_table, caption = "Regression Coefficients", digits = 10)
```

```{r}
variables <- c(
  "food_density",
  "v_CA21_1",
  "v_CA21_449",
  "low_income",
  "food_density:low_income"
  )

# Generate pairwise regression results including interaction
results <- expand_grid(Row = variables, Col = variables) %>%
  mutate(
    Formula = ifelse(
      Row == Col, NA,  # Avoid self-interaction
      paste("crime_density ~", Row, "+", Col)
    )
  ) %>%
  rowwise() %>%
  mutate(
    R_Squared = if (!is.na(Formula)) {
      mod <- lm(as.formula(Formula), data = combined_table)
      summary(mod)$r.squared
    } else {
      NA
    }
  ) %>%
  select(Row, Col, R_Squared)

# Pivot for matrix structure
results_matrix3 <- results %>%
  pivot_wider(names_from = Col, values_from = R_Squared)

# Display the table
kable(results_matrix3, 
      caption = "Regression R^2 Matrix Including Interaction Term",
      digits = 3) %>%
  kable_styling(full_width = FALSE, bootstrap_options = c("striped", "hover", "condensed"))
```
